The Concurrent Calculi Formalisation Benchmark

### 2024-02-01 short-paper/short-paper.tex
comments to sec 1

\begin{abstract}
  POPLMark and POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising
  message-passing concurrency using process calculi.\luca{I like Daniel's proposal below} \dz{Maybe something like, ``\ldots that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.''?} Our benchmark problems
  address three key topics: linearity, scope extrusion, and coinductive
  reasoning.  The goal of this new challenge is to clarify, compare, and advance
  the state of the art, fostering the adoption of proof mechanisation in future
  research on message-passing concurrency.

@@

\section{Introduction}
By establishing a common ground for systematically evaluating mechanisation techniques, 
the POPLMark challenge~\cite{POPLMark} accelerated progress in 
the development of proof assistants, libraries, and best practices. In doing so, it 
spearheaded a shift in the programming language research community towards 
publications that include mechanised proofs. 
Its follow-up challenge, POPLMark Reloaded~\cite{POPLMarkReloaded}, 
facilitated the development of best practices and tool support for proofs 
using logical relations. These challenges were deliberately limited in scope, 
while their authors called for future investigations into such areas as 
coinduction and linear environments. In this spirit, we propose a 
new collection of benchmark problems specifically designed to address typical 
issues that arise when mechanising message-passing concurrency using process 
calculi.

Several papers in this realm already include mechanised proofs
(\eg~\cite{Castro-Perez2021,Castro2020,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), or discuss proof mechanisation techniques
(\eg~\cite{Bengtson2016, Gay2001,
  Thiemann2019, DBLP:conf/forte/ZalakainD21}).  Our
experience, however, is that choosing the appropriate proof mechanisation techniques and 
tools remains a significant challenge with a limited understanding of their 
respective tradeoffs.  This often leads researchers toward a trial-and-error approach, 
resulting in sub-optimal and ad-hoc solutions. In turn, this increases the overall effort 
required for proof mechanisation, and hampers the reuse of techniques and results.
%
It is therefore our view that the research community in message-passing
concurrency (and process calculi, more specifically) will benefit from a standardised set
of challenge problems.

To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties 
when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, 
and \emph{coinductive reasoning} for infinite behaviours. These are certainly not the only factors that may cause issues, but they
are fundamental to concurrency theory, and emerge in most mechanisations.
The mechanisation of novel research results often requires addressing more than
just one of these aspects at once; we see the combination of techniques
as a next step beyond the scope of this challenge (we discuss this further in
\Cref{sec:going-beyond}).
% Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate, evaluate, and compare the many tools and techniques available.
% While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.

Like the authors of the POPLMark challenges, we seek to
answer several questions:
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a session type system, which allows us to introduce linear
reasoning with few definitions. Accordingly, we have chosen not to
include channel delegation and to use a reduction instead of a
transition system semantics.
Linearity is of course interesting in many other subfields of programming languages, but we have chosen to base our challenge on a session type system primarily due to the common use of session types in concurrency theory.
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we use a syntax where
restriction binds two names. This presentation makes the duality
of the two names obvious in the type system.
We have chosen a simple notion of well-formedness for our challenge,
and do not, for instance, consider deadlocks. While more
sophisticated notions of well-formedness are interesting, the
proofs of these properties would complicate the challenge, distracting from the linear aspects.

@@

\alb{Some introductory text to lead to the literature review. \sout{We may want to move this after the next Section, if we want to mirror the 3 challnges}}
\dz{Agreed, I moved it for now.} \\
\dz{I changed the name of this section to ``Existing approaches'' from ``Motivation''. Thoughts?}

>>>>

\begin{abstract}
  %% AM: implemented DZ
% when mechanising  message-passing concurrency using process calculi.

  POPLMark and POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems
  focused on the difficulties that typically arise when mechanising
  formal models of concurrent and distributed programming languages,
  such as process calculi.  Our benchmark problems address three key
  topics: linearity, scope extrusion, and coinductive reasoning.  The
  goal of this new challenge is to clarify, compare, and advance the
  state of the art, fostering the adoption of proof mechanisation in
  future research on message-passing concurrency.

@@

\section{Introduction}
By establishing a common ground for systematically evaluating mechanisation techniques, 
the POPLMark challenge~\cite{POPLMark} accelerated progress in 
the development of proof assistants, libraries, and best practices. In doing so, it 
spearheaded a shift in the programming language research community towards 
publications that include mechanised proofs. 
Its follow-up challenge, POPLMark Reloaded~\cite{POPLMarkReloaded}, 
% facilitated the development of best practices and tool support
aimed to run the same program for proofs 
using logical relations. These challenges were deliberately limited in scope, 
while their authors called for future investigations into such areas as 
coinduction and linear environments. In this spirit, we propose a 
new collection of benchmark problems specifically designed to address typical 
issues that arise when mechanising message-passing concurrency using process 
calculi.

\alb{Is there a sharp difference among those papers?}
Several papers in this realm already include mechanised proofs
(\eg~\cite{Castro-Perez2021,Castro2020,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), or discuss proof mechanisation techniques
(\eg~\cite{Bengtson2016, Gay2001,
  Thiemann2019, DBLP:conf/forte/ZalakainD21}).  Our
experience, however, is that choosing the appropriate proof mechanisation techniques and 
tools remains a significant challenge with a limited understanding of their 
respective trade-offs.  This often leads researchers toward a trial-and-error approach, 
resulting in sub-optimal and ad-hoc solutions. In turn, this increases the overall effort 
required for proof mechanisation, and hampers the reuse of techniques and results.
%
It is therefore our view that the research community in message-passing
concurrency (and process calculi, more specifically) will benefit from a standardised set
of challenge problems.

To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties 
when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, 
and \emph{coinductive reasoning} for infinite behaviours. These are certainly not the only factors that may cause issues, but they
are fundamental to concurrency theory, and emerge in most mechanisations.
The mechanisation of novel research results often requires addressing more than
just one of these aspects at once; we see the combination of techniques
as a next step beyond the scope of this challenge, as  discussed  further in
\Cref{sec:going-beyond}.
% Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate, evaluate, and compare the many tools and techniques available.
% While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.

Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a session type system, which allows us to introduce linear
reasoning with few definitions. Accordingly, we have chosen not to
include channel delegation and to use a reduction instead of a
transition system semantics.
Linearity is of course interesting in many other sub-fields of programming languages, but we have chosen to base our challenge on a session type system primarily due to the common use of session types in concurrency theory.
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we use a syntax where
restriction binds two names. This presentation makes the duality
of the two names obvious in the type system.
We have chosen a simple notion of well-formedness for our challenge,
and do not, for instance, consider deadlocks. While more
sophisticated notions of well-formedness are interesting, the
proofs of these properties would complicate the challenge, distracting from the linear aspects.

@@

\alb{Some introductory text to lead to the literature review. \sout{We may want to move this after the next Section, if we want to mirror the 3 challenges}}
\dz{Agreed, I moved it for now.} \\
\dz{I changed the name of this section to ``Existing approaches'' from ``Motivation''. Thoughts?}
